%!TEX root=../tax-democracy-held.tex

method, notes about q

I don'T think I do the kind of prismatic etc. balanced block design as per \cite[30][]{Brown1980} because I don't have ex-ante theories of what the pattern of understandings will be; I just hypothesize that they will change and straighten up.
As a result, an ANOVA is probably not appropriate (ibid.)

Or maybe I can do this, kinda like \cite[35]{Brown1980} on (Robert?) Reich:

also compare 60 in \cite[35]{Brown1980} that might be it

somewhat oddly, thompson refers to beliefs as ``biases'', as though inbiased facts were a possibility even though operant subjectivity can, by definition not be, well, objective and in operant subjectivity on economic matters in particular, is unlikely to arrive at such uncontentious statements.

summary graph on page 69 is very important.

note in a later conclusion that this should be deveoped a) into a broader standard for deliberative effects and b) into a broader study concerning moral development, cognition and thinking about the economy.

Notice that a lot of the ecologically occuring statements on tax are empty, false or non-sensical and thus should \emph{not} be in the sample.

Shalom Shwartz wrote a q-diss!
	Schwartz, S. H. 1978. Reflections on a Q dissertation and its opposition. Operant
Subjectivity. 1: 78-84.

add lord of the flies to reading list

sociology and q methodology?
q methodology bibliography
q method and econ

notice that as per Brown around 75, q methodology might also be used as a tool for deliberators to figure out exactly what kinds of policies they might agree upon!

Item ``taxation is the price we pay for a civilized society'' (o.W. holmes)


Main Effects		Levels
Consciousness 		a) Con I  	b) con II 	c) con III
Values  			d) personal 	e)lega-political 	f) socio-economic

Q-methodology and habermas! Must look into this!

___
McKeown

notice that factor loadings are actually *not* conventionally very important in Q; mostly factor scores. That'd be different for me; I'd wager the most general hypothesis that factor loadings would improve.

run interviews to gather the q concourse?

discuss the whole adapted vs. naturalistic controversy: in some instances, naturalistic items might work; but not for tax, because here some things are just *never* discussed, that's the very idea of this research.
(it is, in fact, questionable, whether the same problem might not also plague other research, as on love. Non-existent counterfactuals are always important – say, on love. You always have to be careful with hegomony in the language. Of course, the balanced-block design takes care of this already)
In my research, I'll make sure that items are readily comprehensible in lifeworld-terms, which, luckily *is* very possible for taxation, because it is about the oikos.
Of course, non-naturalistic, ``adapted'' items bring back the issue of external observation ,or rather, of things that the researcher has put in here. This problem cannot be entirely negated, but q methodology reduces it significantly.
For once, people can still sort items they do not understand or find odd in the middle, near 0, what Stephenson described as the quantum origin of the scale, with 0 information.
But also, no matter which other items were added to the q-sort, or on which other p-sample it would be replicated, the results of this q-sort would still be valid, because it's about operant behavior.

got to look at the dryzek political discourse q thing from 1993

apparently positive or negative can also be a line in the balanced block approach, as in \citep[23]{McKeown2013}

should I describe severeal conditions of instruction, like: what the actual tax system is like, what it should be like? (Maybe not, because that would be true/false)

In q and p sort, the criterion is saturation.

make people sort items on equity and effifiency? That might make some sense (even though my point is that both expand, which really won't work here.)

---

Niemeyer Diss
- check strong democracy barber

Look at dryzek and list 2001 on metaconsensus.

maybe in opposition to niemeyer, I think you can impose one normative theory – rawls, because it's meta.
just like intersubjective rationality.

Dryzek 1990 discourse theory – this seems important, it is what niemeyer is drawing on!

Intersubective rationality -- that people want the same preferences, because of the same values and beliefs –-- is kind of a strong assumption, because we don't strictly speaking know that such a consistency exists (except and in so far as it merely follows from vNM). But we can accept this as a regulative ideal; something to be strived for, not achieved. And if we never get any further to it, or can't, with that would wither the plausibility of deliberative democracy as an utopia, which, after all, depends on this.
Look into vNM to say that we have collective preferences might imply that we also have this kind of consistency.

get pelletier 1999 as example on q in deliberation

get dryzek 1993

discuss deliberative failure as in niemeyer diss6 69f

i should have some kind of stratified sample, too -- family, entrpreneur, worker etc.

tax attorneys, steuerberater?
bund der steuerzahler?
georgist?
pct?
other experts?

discuss how different tax is from the bloomfield track issue; there's just no immediate way to experience tax. It's all abstraction.
It is, in other worlds, inherently system, not lifeworld.

maybe visit the Finanzamt?
visit the tax business of a firm?

ask people which witnesses they thought were most interesting, influential.

don't just to experts, do practicioners, too.

the problem with practicioners is that there is currently no one out there who does these things, worries about them.

OECD tax people maybe?

my tax items will, obviously, still be real and part of concourse – they are just on the very periphery, which is exactly what I am looking at here.

niemeyer diss 158 compares q methodology to making a species survey of a small area in ecology; that's a good analogy why small numbers work. Also refer here to the brown example of calculating the species factor between arm, leg etc length.

remind people when they do the q-sort that they are not at all being tested; not the correctness, nor the consistency of their ranks (between the two points in time) is tested.

need to argue carefully why the actual policies (taxes) should be q-items, too
1) because they're difficult to understand, as opposed to the bloomfield options
2) because q measures what we'd need to know about them anyway.
3) they are many-dimensional, and their vary dimensionality (much like the other q-items) is of interest. They can't be graded on a n access to non-access dimension as bloomfield (214)

check again dryzek and list 2001 for preference structuration

---

\cite{Dryzek1993} is very important, completely a fan.

The most radical meta-standard that could be formulated would indeed be pure q methodology, with a focus on the loadings and numbers of factors, maybe the three-step of dryzek and niemeyer. That's a tall order, and I'm not sure it works, but that would be the ideal.

Maybe it is not an accident that Dryzek stressed so much the reconstructive merit of q (for democratic theory) and went on to be a formative voice for deliberative democracy.

\cite[52]{Dryzek1993} write: ``Toput it another way, our units of analysis, when it comes to gener- alization, are not individuals but discourses. And we soon get to a point where adding individuals to a study does not yield any new informationunless the extra individuals are truly different-another reason for stressing diversity in subject selection.''
In other words, q methodology works with small n, because it is concerned with the dimensionality, or structure of subjectivities, not the prevalence. Notably, q methodology is downward compatible; if you have a lot of people do it, you can become ever more confident in the relative factor *loadings* (which is what you'd be interested, then). Notably, this is not the case the other way around; if you administater an r survey with *lots* of items, it's still not same as q, because it lacks the same unit.

\cite{Niemeyer2013} is also about a local issue, and even shorter (the birdge thing)

also note that focusing only on the overall q loadings would be sort of against the theory; and conflict with judgemental rotation. Still that would be the general effect of which this might be a special case, if the list/dryzek piece has it right.

also notice that

---
notice that \cite{Brown-2006} does something quite dissimilar; the locals are right, always. I can't do that for tax.
But I should also point out that my research interest is quite circumspect.

--

notice that q methodology does sit a little uneasily with critical theory, including habermas, because it (initially) seems to permit *no* false consciousnesses.
This, of course, is because of the items in the q sort are not supposed to be factual questions, but one of preference (ideology being facts wrapped up in norms.) (uh-uh, this is bullshit)

my selections of q sorts is essentially statements that follow habermas' rules for communicative rationality; it would be non-sensical to ask the for the subejctivity of powered speech.
This might at first seem interesting (and in line with my research), but it violates the central tenet of q methodology, namely to take operant subjectivity serious and place it front and center.
the participant, in other words, is always right.
Also, ``testing'' for powered speech, would be hypothesis testing – which is not supposed to happen with q; I would, in effect, be interested in the movement of individual items.
Instead, I am looking at the dimensions.
Also, if there is such a thing as communicative rationality, it should reveal itself in neatly aligned factors, if not before, than after deliberation.
The same logic still holds; If, as predicted, the factor space clears up, something must have prevented it from clearing in the first place. That something is powered speech.

---

\cite{dryzek2005handle} This seems important, right down my alley for why I need q.

describes why in-depth, etc. are all bad; also how I would become a deliberator in an in-depth interview.

--
\cite{Durning1999} cite this guy; that's a pretty damning review at the beginning of positivism.

--
\cite{hutchinson2012} on complexity, or a post-positivist use and advocacy of q; that's not really what I'm doing, I think. Or maybe it is.

--
\cite{Niemeyer-2011-bookchap}

notice that I am doing something new; there are no q sorts on the economy, or on such highly structured/complex material, or with this research interest.
So I have to develop them myself, from the literature. That's hard.

Should I ask directly for metaconsensus? (no that would be hypothesis testing!) Instead, metaconsensus should emerge as factors.

Notice that single-peaked preference orderings (as in 19 in the above) are a special case that may or may not be observed here; it is difficult to find in terms of taxation, which is multi-dimensional (at least equity and efficiency).

My beef is not with a meta-substanstive standard, that is ok – and there *might* be such a thing as a universally acceptable, rational decision; but it's wrong to assume that in the first place. If might be a regulative ideal, and so should the test be designed. It's something that can be approached, or maybe not, or maybe not in our lifetime. (because what young et al are really worried about is power that persists, and clearly, that might still be the case, and that might be a good argument for deliberation to the extreme in our time (see archong fung on this))

Note how q fits well with my earlier admonition of piercean pragmaticism.

Also note how the standard to which items are held, is again, fittingly, a regulative ideal ``something that people can reasonably disagree with'' – and I might fail that standard.
Importantly, I can still test that.
Also note:
- q items can be double-barrelled etc.
- q items should not be true/false
- q items should not be phrased as hypothesis-testing
- q items should allow deliberators to say anything that could be said (cover the floor with carpet tiles)
- q items should be in terms of the lifeworld
- q items should not ``cheat'' or ``test'' the sorter.

Think again, hard, on why I can't have (too much) stuff from the newspaper in here.

Q methodology also lends itself because, except for the apriori theorized work by McCaffery et al, there isn't an awful lot of theory that would tell us exactly how people ``mis''construe tax.

Factors from q are ideal types of people who score things similarly, the factor SCORE is than the rating a person epitomizing this factor (not part of the sample) would give to that factor.
Factor loadings

Explain again factor loadings vs factor scores

Factor loadings are how much the variance of any particular person is explained by the factor.

Remember that the movement of *individual* items is not meaningful (that goes back to R methodology)

Argue why/if actual taxes should be part of the q sort, too.

Honestly, there is also the point of small n, which is great.

Q methodology (or rather niemeyer/dryzek) should also isolate, to some extent, the charge that I might simply talk people *into* any one tax.

make clear; ideally, this methodology is an *iterative* one, because there might be reasons or voices that the q-sort does not include, but that should be included in future.
So I will document whatever flows out of this, revise the q-sort afterwards.

Also make really clear at the outset that the q-sort is not that all-important; there are gradations.

Also notice that I have tested and expanded on this q-set initially with some items from quite extensive pre-tests, literature research, as well as online.

How are Intersubjective rationality and metaconsensus explicated in Q results?
Figure this out.

Explicate where this research is pushing the boundary:
1) highly policized but structured issue
2) radical opening of tax
3) new application of q method to the subjectivity on complex stuff (cite the article)
4) extension, application of metaconsensus and intersubjective rat. to a complex, technical issue with multidimensional preferences (schedule and base)

Somehow discuss franzis issue about between and within group variance!

Niemeyer 2014 shows exactly this; how it works as a regulative ideal, how there might be differences in how different formats achieve or fail by this goal.

Notice that there might be nothing wrong per se with universal, rational consensus.
After all, what are we do to, but rely on reasons; and reasons always imply (as Habermas wrote) sort of as their telos universal consensus, they are proto-forms of such universal consensus.
It should also be clear that any argument will at the end rest on this and there's no easy formal way around this; I can't math myself out of this.
Science, for example, should work on the basis of reasons alone, and we have a sense that this works, more or less (if only we could get rid of some of the power).
Still I should notive that intersubjective rationality (IR) and metaconsensus (MC) are kind of railings on which to hold; they help constrain interpretation and abstract away some of the problems, but not all.

Notice that absent such a meta-standard, and (the probable) absence of universal consensus (except where required by the format as in consensus conferences) we won't know which consensus is the rational one, if there are several competing ones, again falling back on a substantive struggle.
That's bad.
Also notice that as Niemeyer rights, with universal consensus (which can be measured) it's hard to tell whether the consensus is actually rational, consensus (or dissens), or any chance in preferences migh happen for a number of reasons, including may worrisome non-deliberative ones such as groupthink, polarization, mere exposure/priming or social influecne/desirability.

let people pick pseudonyms.

I am not sure I can even respond to the difference criticism of democracy; I disagree with that.
Point to the article on that, on policy sciences.

cite the pornography example: this is a very good example, and I should cite this to illustrate why it is so important to point out, and explain, missing counterfactuals such as the PCT (much like porn for women, by women).
Same for the q-sort.

make a 2-2 table to explain the choice of data gathering format; one is predefined hypothetico-deductive, the other is unconstrained, qualitative.
And then there is a third ... Q

---

\cite{Ockwell2008}

look at the \cite{Dryzek1993} matrix, maybe this works better than merely beliefs, preferences, values?

---

\cite{Pelletier1999} might be the first guy to actually use q to measure deliberative results.

\cite{price1968} uses q methodology concernign the financial behavior of families, but unfortuantely is quiet on the actual items.

---

I should notice some peculiarities about q, including (but can't do an all-on theory of inquiry, for that, see Brown and Stephenson)
- operant subjectivity
- concourse
praise the opportunity and oppeness and respect for participants this method brings; without being unconstrained.



look again at the bal
