--- 
title: "Give-and-Take"
subtitle: "What citizens would think about taxation, if they could."
description: "Drafts for Maximilian Held's dissertation on taxation and democracy."
author:
  - name: "Maximilian Held"
    affiliation: "Bremen International Graduate School of Social Sciences (BIGSSS)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
bibliography: [held-library.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
github-repo: maxheld83/schumpermas
url: https\://schumpermas.maxheld.de
classoption:
  - a4paper
  - twoside
  - openright
geometry: margin=1.5in
documentclass: memoir
lang: en-US
---

<!-- this includes the travis badge, which does not work in latex -->
```{r, include = knitr:::is_html_output(), eval=knitr:::is_html_output(), child="README.md"}
```

```{r setup, cache = FALSE, include = FALSE}
source("setup.R")
```

<!-- TODO make this an opening page -->

> The spirit of a people, its cultural level, its social structure, the deeds its polity may prepare -- all this and more is written in its fiscal history, stripped of all phrases.
> He who knows how to listen to its message here discerns the **thunder of world history** more clearly than anywhere else.  
> ––- Joseph @Schumpeter [1918: 6]

> The unforced force of the better argument: [@Habermas1996, 305]
> The speaker must choose a comprehensible expression so that speaker and hearer can understand one another. [@Habermas1976: 2f]  
> [A]nyone acting communicatively must, in performing any speech act, raise universal validity claims and suppose that they can be vindicated. [@Habermas1979: 2]  
> Reaching understanding, (...) **the inherent telos of human speech**.
> –– Jürgen Habermas (1987: 287)
<!-- TODO last citation isn't quite right -->

<!-- ![Citizens preparing the press conference at the 2014 CiviCon Citizen -->
<!-- Conference](https://dl.dropboxusercontent.com/u/5341489/images/civicon-prep-poster_small.jpg) -->



## Abstracts {-}

### English {-}

<!-- abstracts are nicer in their own files, hence the weird importing -->
```{r, child="abstract_en.rmd"}
```

### German {-}

```{r, child="abstract_de.rmd"}
```

## Colophon {-}

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/).
The book is available in several formats at [`https://schumpermas.maxheld.de`](https://schumpermas.maxheld.de).
The complete source of the book can be found on [github](https://github.com/maxheld83/schumpermas).
The website is updated after every commit by [travis-ci](https://travis-ci.com/).


### Literate Programming {-}

> Let us change our traditional attitude to the construction of programs:
> Instead of imagining that our main task is to instruct a computer what to do, let us concentrate rather on explaining to humans what we want the computer to do.
>
> --- @Knuth-1984

Most importantly, using `R`, every step from raw data to final factor interpretation can be traced back, using publicly available and vetted code, down to base functions.
<!-- TODO MCH find nice open source quote -->

Machine-readable code, its results, and human-readable explanation of *what that code does* are often written and stored separately.
Not only does such separation invite mistakes when code, result and explanation diverge, but it also makes research harder to reproduce and is conceptually flawed.
<!-- TODO maybe relate to the Rogoff Reinhart paper, cite LaCour retractions -->
At least in the context of statistical analysis, what we want a machine to do, why we want it, and to what effect are *one* intellectual operation, and should be presented as such.
Programs must not be considered black boxes that "do" things of their own accord, to be explained ex-post, but code should be a near equivalent of the same thought expressed in prose.

Donald Knuth has formulated this central tenet of his *Literate Programming* approach thus:

I try to follow this approach by interweaving prose and code in one document, using the [`knitr`](http://yihui.name/knitr/) R package [@knitr].
Aside from published R packages, all code is run *inside* this document upon rendering.
<!-- TODO add knitr, literate programming ref -->
Code will usually not be reproduced in print, but can always be inspected in the [source of this document](https://github.com/maxheld83/schumpermas), "*underneath*" the respective operation or result.

A lot of Q research is funded mostly by tax money, all of this scientific work should be conveniently available to the public.
Research on deliberation and citizen participation, especially, should be *Open* Science.
