--- 
title: "Give-and-Take"
subtitle: "What citizens would think about taxation, if they could."
description: "Drafts for Maximilian Held's dissertation on taxation and democracy."
author:
  - name: "Maximilian Held"
    affiliation: "Bremen International Graduate School of Social Sciences (BIGSSS)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
bibliography: [held-library.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
github-repo: maxheld83/schumpermas
url: https\://schumpermas.maxheld.de
classoption:
  - a4paper
  - twoside
  - openright
geometry: margin=1.5in
documentclass: memoir
lang: en-US
---

<!-- this includes the travis badge, which does not work in latex -->
```{r, include = knitr:::is_html_output(), eval=knitr:::is_html_output(), child="README.md"}
```

```{r setup, cache = FALSE, include = FALSE}
source("setup.R")
```

<!-- TODO make this an opening page -->

> The spirit of a people, its cultural level, its social structure, the deeds its polity may prepare --- all this and more is written in its fiscal history, stripped of all phrases.
> He who knows how to listen to its message here discerns the **thunder of world history** more clearly than anywhere else.
>
> -- Joseph @Schumpeter [1918: 6, emphasis added]

> The unforced force of the better argument: [@Habermas1996, 305]
> The speaker must choose a comprehensible expression so that speaker and hearer can understand one another. [@Habermas1976: 2f]  
> [A]nyone acting communicatively must, in performing any speech act, raise universal validity claims and suppose that they can be vindicated. [@Habermas1979: 2]
> Reaching understanding, (...) **the inherent telos of human speech**.
> [@Habermas-1984: 287]

<!-- ![Citizens preparing the press conference at the 2014 CiviCon Citizen -->
<!-- Conference](https://dl.dropboxusercontent.com/u/5341489/images/civicon-prep-poster_small.jpg) -->


# Abstracts {-}


## English {-}

<!-- abstracts are nicer in their own files, hence the weird importing -->
```{r, child="abstract_en.rmd"}
```


## German {-}

```{r, child="abstract_de.rmd"}
```


# Colophon {-}

> Free as in freedom, not free beer.
>
> --- Richard Stallman, [Free Software Foundation](https://www.fsf.org)

This book was written using **free and open source software (FOSS)**.
This is a [bookdown](http://bookdown.org/) book, written inside [RStudio](http://www.rstudio.com/ide/) and [Atom](http://www.atom.io).
The book is available in several formats at [`https://maxheld.de/schumpermas`](https://maxheld.de/schumpermas).
The complete source of the book can be found on [github](https://github.com/maxheld83/schumpermas).
The website is updated after every commit by [travis-ci](https://travis-ci.com/).

The [rmarkdown](https://rmarkdown.rstudio.com) source of this book interweaves code and prose, following Donald @Knuth-1984's suggestion for **Literate Programming**.
The [R](https://www.r-project.org) code will usually not be reproduced in print, but can always be inspected in the [source of this document](https://github.com/maxheld83/schumpermas), *underneath* the respective operation or result.

This research project strives towards the ideals of **open and reproducible science**.
All results can be reproduced from the raw data using only the scripts contained herein.
<!-- TODO MH add data source -->

<!-- TODO MH add license terms here -->
