# Treatment

<!-- Q2D: A Novel Approach to Measure Deliberative Quality at a Citizen Conference on Taxation -->

<!-- - reports *changes* in factor composition and loadings between pre- and post-sort. -->
<!--    - *factor interpretation* in terms of deliberative democracy (as per Niemeyer: beliefs, values, preferences) -->
<!--    - testing (increased) *meta-consensus and intersubjective rationality* (aka consistency) -->
<!--    - additional *qualitative analyses* of selected, cursorily transcripted discussions from CiviCon-1 -->

<!-- 1. **compare factor loadings**: look at the dimensionality, and loadings of factors befre and after; this would be the broadest kind of result, ideally that the factor space becomes lower dimensional -->
<!-- 2. **pooling** of data (including before and after sorts in *one* dataset for analysis): may show different before/after factors, as well as the change of people -->
<!-- 3. **2nd order analysis** (make factors out of factors): will be limited because of max. 8 cases (factors), but would be very interesting: these might be the *misunderstandings* -->
<!-- 4. **factors of deltas** compute the delta in score for each item (say, item moves from -5 to 2 for a participant), then compute factors of these movements: this may reveal the "misunderstandings", and/or the kinds of learning processes that have happened. -->
<!-- 5. **spiking** of data; place (some of the) "before" factor arrays into the "after" dataset as cases; then do judgmental rotation: this is not a very common method, but may reveal how things have changed from the vantage point of "before". -->

The respectful give and take of deliberative citizen participation is expected to change political views, though in which way, and how to measure it, is less clear. 
On the one hand, conventional survey research offers rigorous and advanced analytics, but is marred by deductive operationalizations and often falls short of a convincing measure of communicative rationality. 
Qualitative approaches, on the other hand, leave much to the researcher's discretion, and risk conflating a measure of deliberative quality with the researcher's substantive theory of justice or reason.

As a mixed method, Q methodology is better suited for the study of deliberative democracy.
It offers a holistic, and operantly defined, yet quantitatively rigorous view on political subjectivity.

<!-- Dryzek and Niemeyer have shown that Q methodology can be a productive *mixed* method to study deliberative democracy: It offers a holistic, and operantly defined, yet quantitatively rigorous view on political subjectivity. -->

<!-- however, q tradittionally provides no clear way to study the *treatment effects* of some participatory intervention -->

Using data gathered before and after the 2014 CiviCon Citizen Conference on taxation, hosted by the author, I suggest several approaches to measure deliberative quality using Q methodology.
At the week-long conference, 16 diverse, self-selected citizens were tasked to design a tax system "from scratch", choosing among possible combinations of base schedule. 
They participated in learning phases, deliberated in moderated small group and plenary sessions, met with experts and held a concluding press conference.
Before and after the conference, citizens sorted 79 statements on taxation and economics according to their subjective viewpoint.

Following traditional Q methodology, sorts were factor analyzed to extract ideal-typical viewpoints shared by participants. 
Before the conference, citizens expressed resentful, radical and moderate viewpoints, including some apparent inconsistencies between beliefs, values and preferences on taxation. 
After the conference, citizens shared decommodifying, pragmatic and critical viewpoints and displayed a simpler, lower-dimensional structuration of viewpoints.

Such results are meaningful and encouraging for the use of Q in deliberation, but fall short of testing a treatment effect, because before- and after-factor models are extracted and rotated separately.
To tease out the effects of deliberation, I therefore also apply a newly developed 2-dimensional Q analysis (Q-2D), based on three-mode Principal Components Analysis (nPCA) by Tucker and Kroonenberg.
On the one hand, results indicate that, after considerable computation and graphical transformation, the effects of deliberation can be meaningfully interpreted using Q2D. 
Participation in the Citizen Conference appears to increase substantive consistency of political viewpoints, seemingly meeting a demanding standard for deliberative democracy. 
On the other hand, a unified longitudinal factor model may also unreasonably restrict the change of subjectivity during deliberation. 
Analyzed separately, there is some evidence that before and after factor models are, in fact, incongruent.

Alternative analytical approaches to Q data in empirical deliberative democracy research are discussed.

Political subjectivity has been conceptualized by Dryzek and Niemeyer to involve *beliefs* and *values*, as well as *preferences*.
These are conceptually *independent dimensions* and the very status of some assertion as either value or belief is often contested, and thereby part of political subjectivity proper.
Here, too, Q2D can significantly improve upon conventionally analyzed, repeated Q studies by yielding *jointly extracted and rotated*, two-dimensional ideal viewpoints vis-a-vis the *same* items, readily interpretable in terms of meta-consensus and intersubjective rationality.

Algorithms and visualizations in open-source R-code are presented, conceptual limitations of Q2D are discussed and compared to alternative analytical approaches to Q data in empirical deliberative democracy research.

<!-- - Communality / nloads of factors seem to change (different people!)
- Eigenvalues increase!
- Factor of factors?
- Factor of item movements?
Look at factor CONGRUENCE (the angles!) -->

<!-- make screeplot with before/after data in one plot, that's essentially what you're hypothesizing -->



```{r diff, eval = FALSE}
diff <- NULL
diff$data <- q_sorts[,,"after"] - q_sorts[,,"before"]

test <- q.corrplot(corr.matrix = cor(diff$data, method = "pearson")) + ggtitle("Correlation of Movements")


diff$howmany <- q.nfactors(dataset = diff$data, cutoff = 8)
diff$howmany$screeplot

diff$howmany.r <- q.nfactors(dataset = t(diff$data))
diff$howmany

diff$results <- qmethod(dataset = diff$data, nfactors = 3, rotation = "equamax", forced = FALSE, cor.method = "pearson", threshold = "none", allow.confounded = TRUE, distribution = diff$distro)

diff$distro <- c(
  -7, 
  -6, 
  rep(-5, 2), 
  rep(-4, 4),
  rep(-3, 6),
  rep(-2, 9),
  rep(-1, 10),
  rep(0, 11),
  rep(1, 10),
  rep(2, 9),
  rep(3, 6),
  rep(4, 4),
  rep(5, 2),
  6,
  7
)

diff$results$f_char$characteristics

q.scoreplot.ord(results = diff$results, factor = 1)

q.corrplot(corr.matrix = (cor(t(diff$data))))
```



```{r threeway, eval = FALSE}
library(ThreeWay)
help(T3)
dimnames(q_sorts)
keyneson$threeway <- T3(data = q_sorts, laba = dimnames(q_sorts)[[1]], labb = dimnames(q_sorts)[[2]], labc = dimnames(q_sorts)[[3]])
```


<!-- is there a two-dimensional equivalent to PCA? -->


<!-- crazy idea for two-dimensional PCA, just make the lower diag different, this is via qlist from Tim Michel:
i J Christopher:

I've been reading all the responses my query  has generated, to reflect on your response:

1) I understood what you meant by fusing the Q-Sorts together.
2) However, I agree with others on the topic that it doesn't make sense to simply fuse the sorts together and then do the calculations as if they were one large Q-Sort
3) I reiterate that using a block diagonal correlation matrix would seem a more reasonable approach

i.e.  [ M   0  ] , where M are the correlations based on the 1st Qsort and N the 2nd.
      [ 0   N  ]

4) That would get away from correlating apples and oranges so to speak
5) This way we would get a cleaner result and the calculated factors would possibly reflect the subject positions on Qsort1 and Qsort2 without the other noise in the data.
6) I think one would still have to make a case that the qsets are related and would have to administered in the same temporal frame.

Now in my limited exposure to Q-Methodology, it seems to me that it is employed to find clusters of opinions or subjectivities on the topic outlined int the Q-sets. I would think that combining the Q-Sorts in this manner would identify clusters of opinions & subjectivities in the combined Q-Sets.
 -->
